---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: FIRST LAST
---

# A. Introduction  
This file contains the **instructions** for the final project.  

To get you setup, you will need to:  

  - Student #1: create a repository on your GitHub account. You can call this repository "2026dsa_finalproject_groupX", **where X is the number of your group**. Make it public, add a README, add a .gitignore for R.  
  - Student #1: follow the steps we did in class to start a new RStudio project with version control.  
  - Student #1: in your computer, create the sub-folders code, data, output, and move your data set into the `data` folder. Also, student 1 moves this current script into the `code` folder. Do a git stage, commit, push.  
  - Student #1: on GitHub, go the repository settings and invite your partner to be a collaborator in the repository. That will give them push permission.  
  - Now, both students should clone this repository on their computers like we did in class. Make sure this step works well and that you can pull and push from GitHub.  
  - Student 2, after cloning, does a git pull to get all these updates on their computer.  
  - Student 1 and 2 work together to update the README file. README files should explain what the repository is about, the goals of that project, who is working in it, and any other important details you may find.  

# B. Defining the problem  

You are being provided with a **corn variety trial** data set.  
This data contains over **164,000 rows** spanning **10 years** (2014-2023), **45 sites** across the USA, for a total of **270** site-years, where over **5,000** corn hybrids were evaluated.  

You are being provided with the following **training data** for years 2014-2023:   

  - Trait information (including site, year, hybrid, yield, and grain moisture)  
  - Meta information (including site, year, previous crop, longitude, and latitude)  
  - Soil information (including site, year, soil pH, soil organic matter, soil P, soil K)  

You are also being provided with the following **testing data** for year 2024:  
  - Submission information (site, year, hybrid, **no yield**)  
  - Meta information (same as training)  
  - Soil information (same as training)  

> You are tasked in training machine learning models using the training data to predict yield on the test data.  

One difference from class is that **you are not provided with the test data yield (predicted variable)**, only the predictor variables.  
Therefore, using the training data wisely to create predictive models will be important.  

## 1. Cleaning the raw datasets
Loading necessary libraries

```{r}
library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
```
Loading training and testing data
  
```{r}
training_trait <- read_csv("../data/training/training_trait.csv")
training_meta <- read_csv("../data/training/training_meta.csv")
training_soil <- read_csv("../data/training/training_soil.csv")
testing_submission <- read_csv("../data/testing/testing_submission.csv")
testing_meta <- read_csv("../data/testing/testing_meta.csv")
testing_soil <- read_csv("../data/testing/testing_soil.csv")
```

```{r}
training_trait %>% distinct(site)
training_meta %>% distinct(site)
training_soil %>% distinct(site)
testing_submission %>% distinct(site)
testing_meta %>% distinct(site)
testing_soil %>% distinct(site)
```

Inspecting the training_trait dataset

```{r}
head(training_trait)
names(training_trait)

#Summary of the variables in the data
summary(training_trait)

# Number of distinct sites
training_trait %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_trait))

# Checking for duplicated rows
dup_rows_trait <- training_trait[duplicated(training_trait), ]
dup_all_trait <- training_trait[duplicated(training_trait) | duplicated(training_trait, fromLast = TRUE), ]
dup_all_trait

sum(duplicated(training_trait))

training_trait_clean <- training_trait[!duplicated(training_trait), ]
```

There are 45 unique sites. We can see that the variable date_harvested has 491 missing observations out of 164477. There are 8 duplicated rows in the entire dataset so we removed those duplicated rows. 

```{r}

head(training_meta)
names(training_meta)
```

```{r}
#Summary of the variables in the data
summary(training_meta)

# Number of distinct sites
training_meta %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_meta))
```


```{r}
table(training_meta$year, training_meta$site)
dup_rows_meta <- training_meta[duplicated(training_meta), ]
dup_rows_meta
dup_all_meta <- training_meta[duplicated(training_meta) | duplicated(training_meta, fromLast = TRUE), ]
dup_all_meta
sum(duplicated(training_meta))
```

The are 43 unique sites in training_meta dataset. And the variable previous_crop has 24 missing observations out of 256. 

Some sites have duplicate observations. Each site year should have one row. We checked and found that following sites have duplicate observations.

IAH1: 3 obs in 2014 (Duplicate = 2)
MOH1: 2 obs in 2020 (Duplicate = 1)
TXH1: 3 obs in 2017 (Duplicate = 2)

Also, MOH1-rep1 and MOH1-rep 2 have same information. Can merge them as MOH1 and keep only 1? No because in training_soil we have seperate observations for MOH1_1 and MOH1_2. So, for uniformness we can rename MOH1_rep1 and MOH1_rep 2 to MOH1_1 and MOH1_2 respectively. 

TXH1-Dry, TXH1-EArly and TXH1-Late have same values. Can we just take it as TXH1 and keep 1 observation for 1 site-year? Answer: Yes because they have the same latitude and longitude but once we get the daymet merged we should have all three entries as the training_trait has spearate observation for TXH1-Dry, TXH1-EArly and TXH1-Late.

```{r}
training_meta_clean$site[training_meta_clean$site %in% c("TXH1- Dry", "TXH1- Early", "TXH1- Late")] <- "TXH1"
```


Removing the duplicated rows

```{r}
training_meta_clean <- training_meta_clean[!duplicated(training_meta_clean), ] 
```

This did not remove the duplicated row for the site MOH1 in 2020 because the value for both entries are slightly different. But as the values are same to 2 decimal points, we will remove this duplicated row manually.

```{r}
training_meta_clean <- training_meta_clean %>% distinct(year, site, previous_crop, .keep_all = TRUE)
```

The sites MOH1_Rep1 and MOH1_Rep2 in 2018 will be renamed to MOH1_1 and MOH1_2 as it will match the training spoi sites MOH1_1 and MOH1_2 in year 2018.

```{r}
training_meta_clean$site[training_meta_clean$site == "MOH1_Rep1"] <- "MOH1_1"

training_meta_clean$site[training_meta_clean$site == "MOH1_Rep2"] <- "MOH1_2"
```

Let's investigate the missing observations in the previous_crop
```{r}
table(training_meta_clean$previous_crop, useNA = "ifany")
```
Renaming some crops for uniformness
```{r}

training_meta_clean$previous_crop[training_meta_clean$previous_crop %in% c("2019/20 wheat", "wheat", "Winter wheat", "Winter Wheat")] <- "Wheat"

training_meta_clean$previous_crop[training_meta_clean$previous_crop  %in% c("wheat and Double Crop soybean", "wheat/double crop soybean", "wheat/soybean")] <- "Wheat/Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "cereal rye"] <- "Rye"


training_meta_clean$previous_crop[training_meta_clean$previous_crop == "corn"] <- "Corn"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "cotton"] <- "Cotton"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Fallow most of 2014 winter planted in fall of 2014 then sprayed with Glystar 24 floz/a on 5/3/15  and killed spring of 2015 spray"] <- "Fallow"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Lima beans followed by rye cover crop"] <- "Lima/Rye"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "peanut"] <- "Peanut"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Small Grains and Double Crop soybean"] <- "Grains/Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "sorghum"] <- "Sorghum"

training_meta_clean$previous_crop[training_meta_clean$previous_crop %in% c("soybean", "Soybeans")] <- "Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "soybeans with fall cereal rye cover"] <- "Soybean/Rye"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "sugar beet"] <- "Sugarbeet"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "soybean/pumpkin"] <- "Soybean/Pumpkin"

```


```{r}
table(training_meta_clean$previous_crop, useNA = "ifany")
```

```{r}
head(training_soil)
names(training_soil)
```

```{r}
#Summary of the variables in the data
summary(training_soil)

# Number of distinct sites
training_soil %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_soil))

#Looking for duplicated rows
dup_rows_soil <- training_soil[duplicated(training_soil), ]
dup_rows_soil
```

There are no duplicated rows in training_soil dataset.

```{r}
training_soil_clean <- training_soil %>% 
  separate(site, into = c("site", "years"), sep = -5) %>%
  select(-years)

training_soil_clean %>% distinct(site)
```

Training_soil dataset has 40 unique sites. We can also observe that variable soilpH has 2 missing observations and om_pct has 8 missing observations.

## 2. Saving cleaned datasets. (training_trait_clean, training_meta_clean and training_soil_clean)
```{r}
write.csv(training_trait_clean, "../data/training/training_trait_clean.csv", row.names = FALSE)

write.csv(training_meta_clean, "../data/training/training_meta_clean.csv", row.names = FALSE)

write.csv(training_soil_clean, "../data/training/training_soil_clean.csv", row.names = FALSE)
```


# C. Open-source data  
Notice you were provided with the **longitude and latitude** of most of the site-years.  
You are encouraged to use this information to pull any external open-source data sets like **weather**, or anything else you may find useful.  

This will be up to each team.
Part of creating a great model is incorporating different pieces of information.  

# D. Feature engineering  
Your goal is to create good predictor variables in your data set for modelling.  

It is up to each team how you will perform feature engineering.  
For ex. in class we summarized weather on a monthly basis, but perhaps this is not the best approach.  


# E. Modeling strategies  
You are asked to fully train two different machine learning models:  
  - XGboost (everyone)  
  - **One more model of your choice not used in class**.  

You can find a list of all tydymodels supported models here:  
https://www.tidymodels.org/find/parsnip/  

# F. Training strategies  
For both XGboost and your model of choice, the entire training process is up to you.  
That means you decide things like  
  - Data split proportions  
  - Pre-processing steps  
  - Hyper-parameters to fine tune  
  - Search strategy  
  - Resampling strategy  
  - Which metric to use to select best model  
  - Which type of model selection to use (e.g., best model, best within 1 sd, etc.)  
  
# G. Communicating your results  
You will be required to create a **shiny app** to communicate your results.  
Select one of your final models (XGboost or the other one), and display its results in the app.
While the shiny app design is up to you, it should contain:  
  - At least one EDA on yield based on the training data  
  - EDA on at least 1 predictor variable of each group (soil, weather)  
  - At least one interactive component  
  - A plot showing the most important variables in your model  
  - A plot for predicted vs observed with R2 and RMSE on your testing set (not the one I provide)  

# H. Troubleshooting  
You will for sure run into code issues.  
This is common and expected, and part of the learning process.  

While this is an individual project, I do encourage all students to help each other, especially as you will likely run into similar problems.  

For that to work, we will be using **GitHub** to ask and answer questions.  

ALL QUESTIONS should be asked in our course GitHub page (https://github.com/leombastos/2026_dsa) under "Issues". **Please do not use email for asking questions**.

Make sure to **"Watch"** the repository so you get notified when someone posts on Issues.  

> I anticipate all of you will have multiple questions. The goal with using GitHub is that you can help each other. You will be graded for participation both in asking questions on GitHub and also helping others with their questions.  

With that, when you have issues running code, here are a few resources you can use, in chronological order:  

- **Yourself**: Read the error message, see if you can interpret it and understand what is going on. A message like "Error: object yield could not be found" is self-explanatory.    
- **Google**: Sometimes just copying an error message and pasting on Google can help you find posts with the answer.  
- **Peers**: ask your classmates using GitHub.  
- **Me**: after you have gone through all the sources above without success, I will certainly be glad to assist you. I want to be the last resource you use because that's how it will be after our class is finished: I will be available to assist you in anything R-related in your career, but you will also need to attempt solving them before you reach out.  

# I. Turning it in  
You are required to turn in the following:  
  - The github repository URL containing ALL the code for the ML training process  
  - The completed `testing_submission` file with your predictions for the test data that I shared with you. **NOTE: only fill in the NA cells, DO NOT CHANGE anything else! If you send this file with different column names, etc, you WILL LOSE POINTS!**  
  - The URL for the published shiny app  
  - A powerpoint slide named "2026dsa_teamX.ppt" where X is the number of your team. The deck should contain a single slide showing 
  
Submit the github URL, the `testing_submisison` csv, and the shiny app URL by **April 30th** 11:59 pm on the following link:  
https://forms.gle/Es34gk74DHBBSEoa8 
  
# J. Grading  
You will be graded based on:  
  - following general workflows we did in class for all steps (open-source data retrieval and processing, feature engineering, ML model training, etc.)  
  - following reproducible standards like sensible folder and file naming, script organization, script comments, etc.  
  - collaborating with your partner in GitHub  
  - submitting all three pieces of information (2 URLs, 2 files)  
  - submitting on time  
  - having a shiny dashboard with the minimum requirements listed above  

# K. Team work in GitHub  
Whether you are working with your future-self or as duos, make sure to stage, commit, and push after finishing each of the sub-sections above. When committing, write commit messages that are short and descriptive (e.g., finished wrangling).  

If you are working in duos, make sure to split the workload. I will check your collaborative work through the commit history, and if one student has contributed significantly more than the other, than that will impact grades.  

**Tip 1**: to avoid merge conflicts, make sure to **pull** first, and then start working locally. That will ensure that any changes made by your partner will be "downloaded" before you make changes to the files locally.  

**Tip 2**: make use of the Issues on this repository to set up to-do lists and assign tasks to different people. You can also use each issue/task to discuss how things should be run and get to an agreement.  

# L. Extra  
Once each team submits their predictions on my test set, I will compare their predictions with the observed values (that you don't have).  

The team that gets the highest **R2** or lowest **RMSE** will receive an award and bragging rights.  

