---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: FIRST LAST
---

# A. Introduction  
This file contains the **instructions** for the final project.  

To get you setup, you will need to:  

  - Student #1: create a repository on your GitHub account. You can call this repository "2026dsa_finalproject_groupX", **where X is the number of your group**. Make it public, add a README, add a .gitignore for R.  
  - Student #1: follow the steps we did in class to start a new RStudio project with version control.  
  - Student #1: in your computer, create the sub-folders code, data, output, and move your data set into the `data` folder. Also, student 1 moves this current script into the `code` folder. Do a git stage, commit, push.  
  - Student #1: on GitHub, go the repository settings and invite your partner to be a collaborator in the repository. That will give them push permission.  
  - Now, both students should clone this repository on their computers like we did in class. Make sure this step works well and that you can pull and push from GitHub.  
  - Student 2, after cloning, does a git pull to get all these updates on their computer.  
  - Student 1 and 2 work together to update the README file. README files should explain what the repository is about, the goals of that project, who is working in it, and any other important details you may find.  

# B. Defining the problem  

You are being provided with a **corn variety trial** data set.  
This data contains over **164,000 rows** spanning **10 years** (2014-2023), **45 sites** across the USA, for a total of **270** site-years, where over **5,000** corn hybrids were evaluated.  

You are being provided with the following **training data** for years 2014-2023:   

  - Trait information (including site, year, hybrid, yield, and grain moisture)  
  - Meta information (including site, year, previous crop, longitude, and latitude)  
  - Soil information (including site, year, soil pH, soil organic matter, soil P, soil K)  

You are also being provided with the following **testing data** for year 2024:  
  - Submission information (site, year, hybrid, **no yield**)  
  - Meta information (same as training)  
  - Soil information (same as training)  

> You are tasked in training machine learning models using the training data to predict yield on the test data.  

One difference from class is that **you are not provided with the test data yield (predicted variable)**, only the predictor variables.  
Therefore, using the training data wisely to create predictive models will be important.  

## 1. Cleaning the raw datasets
Loading necessary libraries

```{r}
#| message: false
#| warning: false
library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(tidyverse) 
library(USAboundaries) # for US state boundaries
library(sf) # for US map
library(daymetr) # For opendata download
library(lubridate)
library(ggridges)
library(tidymodels)   
library(finetune)     
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   
```

Loading data 

```{r}
training_trait <- read_csv("../data/training/training_trait.csv")
training_meta <- read_csv("../data/training/training_meta.csv")
training_soil <- read_csv("../data/training/training_soil.csv")
testing_submission <- read_csv("../data/testing/testing_submission.csv")
testing_meta <- read_csv("../data/testing/testing_meta.csv")
testing_soil <- read_csv("../data/testing/testing_soil.csv")
```

Unique sites for each dataset

```{r}
training_trait %>% distinct(site)
training_meta %>% distinct(site)
training_soil %>% distinct(site)
testing_submission %>% distinct(site)
testing_meta %>% distinct(site)
testing_soil %>% distinct(site)
```

Inspecting and cleaning the training_trait dataset

```{r}
head(training_trait)
names(training_trait)

#Summary of the variables in the data
summary(training_trait)

# Number of distinct sites
training_trait %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_trait))

# Checking for duplicated rows 
dup_rows_trait <- training_trait[duplicated(training_trait), ]
dup_all_trait <- training_trait[duplicated(training_trait) | duplicated(training_trait, fromLast = TRUE), ]
dup_all_trait

sum(duplicated(training_trait))

training_trait_clean <- training_trait[!duplicated(training_trait), ]

#Creating two columns for site where site will be used for merging with other dataset and site_original will have the original sites for more information

training_trait_clean <- training_trait_clean%>%
  mutate(
    site_original = site,
    site = substr(site, 1, 4)
  )

training_trait_clean %>% distinct(site)

training_trait_clean %>% distinct(site_original)
```

There are 45 unique sites. We can see that the variable date_harvested has 491 missing observations out of 164477. There are 8 duplicated rows in the entire dataset so we removed those duplicated rows. After this, there are 0 duplicated rows in training_trait dataset. 

Cleaning training_meta dataset

```{r}

head(training_meta)
names(training_meta)
```

```{r}
#Summary of the variables in the data
summary(training_meta)

# Number of distinct sites
training_meta %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_meta))
```


```{r}
table(training_meta$year, training_meta$site)
dup_rows_meta <- training_meta[duplicated(training_meta), ]
dup_rows_meta
dup_all_meta <- training_meta[duplicated(training_meta) | duplicated(training_meta, fromLast = TRUE), ]
dup_all_meta
sum(duplicated(training_meta))
```

The are 43 unique sites in training_meta dataset. And the variable previous_crop has 24 missing observations out of 256. 

Some sites have duplicate observations. Each site year should have one row. We checked and found that following sites have duplicate observations.

IAH1: 3 obs in 2014 (Duplicate = 2)
TXH1: 3 obs in 2017 (Duplicate = 2)
MOH1: 2 obs in 2020 (Duplicate = 1)

Also, MOH1-rep1 and MOH1-rep 2 have same information. Can merge them as MOH1 and keep only 1? Yes, because they have pretty close longitude and latitude values in the meta dataset which will have similar weather information. So for meta dataset we can do that but for trait dataset we do not do that.


TXH1-Dry, TXH1-Early and TXH1-Late have same values. Can we just take it as TXH1 and keep 1 observation for 1 site-year? Answer: We can do for meta dataset for getting weather information. But for the trait and soil dataset, we should keep it as it is. Because in they have the same latitude and longitude but once we get the daymet merged we should have all three entries as the training_trait has separate observation for TXH1-Dry, TXH1-EArly and TXH1-Late.

```{r}
training_meta$site[training_meta$site %in% c("TXH1- Dry", "TXH1- Early", "TXH1- Late")] <- "TXH1"

training_meta$site[training_meta$site %in% c("MOH1- rep 1", "MOH1- rep 2")] <- "MOH1"
```


Removing the duplicated rows

```{r}
training_meta_clean <- training_meta[!duplicated(training_meta), ] 
```

This did not remove the duplicated row for the site MOH1 in 2020 because the value for both entries are slightly different. But as the values are same to 2 decimal points, we will remove this duplicated row manually.

```{r}
training_meta_clean <- training_meta_clean %>% distinct(year, site, previous_crop, .keep_all = TRUE)
```


Let's investigate the missing observations in the previous_crop
```{r}
table(training_meta_clean$previous_crop, useNA = "ifany")
```

Renaming some crops for uniformness

```{r}

training_meta_clean$previous_crop[training_meta_clean$previous_crop %in% c("2019/20 wheat", "wheat", "Winter wheat", "Winter Wheat")] <- "Wheat"

training_meta_clean$previous_crop[training_meta_clean$previous_crop  %in% c("wheat and Double Crop soybean", "wheat/double crop soybean", "wheat/soybean")] <- "Wheat/Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "cereal rye"] <- "Rye"


training_meta_clean$previous_crop[training_meta_clean$previous_crop == "corn"] <- "Corn"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "cotton"] <- "Cotton"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Fallow most of 2014 winter planted in fall of 2014 then sprayed with Glystar 24 floz/a on 5/3/15  and killed spring of 2015 spray"] <- "Fallow"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Lima beans followed by rye cover crop"] <- "Lima/Rye"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "peanut"] <- "Peanut"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "Small Grains and Double Crop soybean"] <- "Grains/Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "sorghum"] <- "Sorghum"

training_meta_clean$previous_crop[training_meta_clean$previous_crop %in% c("soybean", "Soybeans")] <- "Soybean"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "soybeans with fall cereal rye cover"] <- "Soybean/Rye"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "sugar beet"] <- "Sugarbeet"

training_meta_clean$previous_crop[training_meta_clean$previous_crop == "soybean/pumpkin"] <- "Soybean/Pumpkin"

```


```{r}
table(training_meta_clean$previous_crop, useNA = "ifany")
```

```{r}
head(training_soil)
names(training_soil)
```

```{r}
#Summary of the variables in the data
summary(training_soil)

# Number of distinct sites
training_soil %>% distinct(site)

#Missing values for variables 
colSums(is.na(training_soil))

#Looking for duplicated rows
dup_rows_soil <- training_soil[duplicated(training_soil), ]
dup_rows_soil
```

There are no duplicated rows in training_soil dataset.

```{r}
training_soil_clean <- training_soil %>% 
  separate(site, into = c("site", "years"), sep = -5) %>%
  select(-years)

training_soil_clean <- training_soil_clean%>%
  mutate(
    site_original = site,
    site = substr(site, 1, 4)
  )

training_trait_clean %>% distinct(site)

```

Training_soil dataset has 40 unique sites. We can also observe that variable soilpH has 2 missing observations and om_pct has 8 missing observations.

## 2. Saving cleaned datasets. (training_trait_clean, training_meta_clean and training_soil_clean)
```{r}
write.csv(training_trait_clean, "../data/training/training_trait_clean.csv", row.names = FALSE)

write.csv(training_meta_clean, "../data/training/training_meta_clean.csv", row.names = FALSE)

write.csv(training_soil_clean, "../data/training/training_soil_clean.csv", row.names = FALSE)
```


# C. Open-source data  
Notice you were provided with the **longitude and latitude** of most of the site-years.  
You are encouraged to use this information to pull any external open-source data sets like **weather**, or anything else you may find useful.  

## 1. Opening clean data with latitude and longitude.
```{r}
training_meta_clean <- read_csv("../data/training/training_meta_clean.csv")

head(training_meta_clean, 5)

training_meta_clean %>% distinct(site)
```

##2. Explore locations distribution in US.
```{r}
states <- us_states() %>%
  filter(!(state_abbr %in% c("PR","AK","HI")))

ggplot() +
  geom_sf(data = states) +
  geom_point(data = training_meta_clean, aes(
             x = longitude,
             y = latitude,
             color = site))
```
There are traits outside US, so lets remove them to download Daymet data properly, since trails experiment is inside US and daymet has good accuracy in US.

```{r}
met_filt <- training_meta_clean %>%
  filter(
    longitude >= -125 & longitude <= -70, 
    latitude >= 25 & latitude <= 50 # This will remove Europe location
  ) # This will remove Canada location, since Lat and Long can match.


summary(met_filt)
```

```{r}
states <- us_states() %>%
  filter(!(state_abbr %in% c("PR","AK","HI")))

ggplot() +
  geom_sf(data = states) +
  geom_point(data = met_filt, aes(
             x = longitude,
             y = latitude,
             color = site))
```

```{r}
met_filt %>% 
  select(year) %>%
  unique()
```


Now , with 37 locations inside US, lets download the open data.
37 locations, 10 years.

## 3. Downloading open data.
```{r}
names(met_filt)

daymet_all <- met_filt %>%
  mutate(weather = pmap(list(.y = year,
                             .site = site,
                             .lat = latitude,
                             .lon = longitude),
                        function(.y, .site, .lat, .lon) 
                         download_daymet(site = .site,
                                    lat = .lat,
                                    lon = .lon,
                                    start = .y,
                                    end = .y,
                                    silent = T,
                                    simplify = T) %>%
                          rename(.year = year,
                                 .site = site)
                        
                        ))

daymet_all$weather[1]

daymet_all_2 <- daymet_all %>% # This will just rename lat and long to avoid complications with merged columns in the future
  rename(lat = latitude,
         long = longitude)

```
```{r}
daymet_all_unnest <- daymet_all_2 %>%
  unnest(weather) %>%
  pivot_wider(names_from = measurement,
              values_from = value) %>%
  clean_names()

daymet_all_unnest 
```

## 4. EDA of downloaded data
```{r}
daymet_all_unnest %>%
  pivot_longer(cols = dayl_s:vp_pa) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~name, scales = "free")
```

## 5. Saving open data.
```{r}
write_csv(daymet_all_unnest,
          file="../data/training/training_meta_weather_clean.csv")
```


This will be up to each team.
Part of creating a great model is incorporating different pieces of information.  

# D. Feature engineering  
Your goal is to create good predictor variables in your data set for modelling.  

## 1. Reading open data
```{r}
fieldweather <- read_csv("../data/training/training_meta_weather_clean.csv")
head(fieldweather, 5)
```

## 2. Creating month info and analyzing basics
```{r}
fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, site, lat, long,
                previous_crop,
                yday,
                dayl.s = dayl_s, 
                prcp.mm = prcp_mm_day,
                srad.wm2 = srad_w_m_2, 
                tmax.c = tmax_deg_c, 
                tmin.c = tmin_deg_c,
                vp.pa = vp_pa
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>%
  mutate( date = as.Date(date_chr, "%Y/%j")) %>%
  # Extracting month from date  
  mutate(month = month(date)) %>%
  mutate(month_abb = month(date, label = T))

names(fe_month)
```

```{r}
fe_month_sum <- fe_month %>%
  
  group_by(year, site, month_abb, previous_crop) %>%
  summarise(across(.cols = c(dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean,
                   .names = "mean_{.col}"),
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}")) %>%
  ungroup()


fe_month_sum
```

```{r}
fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(mean_dayl.s : sum_prcp.mm

) %>%
  mutate(varname = paste0(name, "_", month_abb)) %>%
  dplyr::select(-name, -month_abb) %>%
  pivot_wider(names_from = varname, 
              values_from = value) %>%
  mutate(across(-c(year, site, previous_crop), ~ round(., 1))) #. Represent each one of the columns, ~ is just convention to use across and a function inside.

fe_month_sum_wide 
```

## 4. EDA of feat eng.
```{r}
finalplots <- fe_month_sum %>%
  pivot_longer(mean_dayl.s:sum_prcp.mm
) %>%
  group_by(name) %>%
  nest() %>%
  mutate(plot = map2(data, name,
                     ~ggplot(data = .x,
       aes(y = month_abb,
           x = value,
           fill = stat(x)
           )) +
  geom_density_ridges_gradient(scale = 3,
                               rel_min_height = 0.01) +
  scale_fill_viridis_c(option = "C", alpha = 0.5) +
  theme(legend.position = "none") +
  labs(x = .y)
                       ))
  
finalplots$plot
```

## 5. Exporting Feature Eng. Data
```{r}
write_csv(fe_month_sum_wide,
          "../data/training/training_meta_weather_monthsum.csv")
```

## 6. Opening all data
```{r}

train_trait_dataset <-read_csv("../data/training/training_trait_clean.csv") %>%
  mutate(
    date_planted   = as.Date(date_planted, format = "%m/%d/%Y"),
    date_harvested = as.Date(date_harvested, format = "%m/%d/%Y"))


train_meta_dataset<-read_csv("../data/training/training_meta_weather_monthsum.csv")


train_soil_dataset <-read_csv("../data/training/training_soil_clean.csv")

#Training soil dataset had two observations for MOH1_1 and MOH1_2 in year 2018 and 2020 but in the training trait we do not have that. So we will aggregate the values of MOH1_1 and MOH1_2 into one MOH1 for year 2018 and 2020.

train_soil_dataset <- train_soil_dataset %>%
  group_by(site, year) %>%
  summarise(
    across(c(soilpH, om_pct, soilk_ppm, soilp_ppm), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

train_meta_dataset %>% distinct(year)
train_trait_dataset %>% distinct(year)
train_soil_dataset %>% distinct(year)
```

## 7 Aggregating the training_trait dataset to year-site level.
```{r}
train_trait_aggregated <- train_trait_dataset %>%
  group_by(year, site, hybrid, site_original) %>%
  summarise(
    yield_mg_ha = mean(yield_mg_ha, na.rm = TRUE),
    n_plots = n(),
    .groups = "drop"
  )

head(train_trait_aggregated)
```

## 8. Joining data
```{r}
train_partial_dataset <- train_trait_aggregated %>%
  left_join(train_meta_dataset, by = c("site", "year"))

train_full_dataset <- train_partial_dataset %>%
  left_join(train_soil_dataset, by = c("site", "year"))

```

## 9. Exporting Final Data
```{r}
write.csv(train_full_dataset,
          file = "../data/training/train_full_dataset.csv")
```


## Final data with features engeneered and all datasets merged.

# Planting months were from March to June
# Harvesting months were from August to December

It is up to each team how you will perform feature engineering.  
For ex. in class we summarized weather on a monthly basis, but perhaps this is not the best approach.  

##10 Final dataset inital checks for missingness and summaries 
```{r}
train_full_dataset <- read_csv("../data/training/train_full_dataset.csv")
```

```{r}
colSums(is.na(train_full_dataset))
```
We see that for exactly 6127 observations, weather variables have missing observations. 

```{r}
#Site and years that have missing observations for weather variables
train_full_dataset %>%
  filter(is.na(mean_dayl.s_Jan)) %>%
  distinct(year, site) %>%
  arrange(year, site)
```
These are the site-years that werenot included in the training_meta dataset.

Since all weather variables are absent we will drop those rows.

```{r}
train_model_dataset <- train_full_dataset %>%
  filter(!is.na(mean_dayl.s_Jan)) 

train_model_dataset <- train_model_dataset %>%
  mutate(
    previous_crop = if_else(is.na(previous_crop), "Unknown", previous_crop)
  )

colSums(is.na(train_model_dataset))

```
We can still see missing observation in soil-related variables 
The dataset training_soil_dataset only has observations for year 2015 to 2023 but the meta and trait dataset has observations from 2014 to 2023. This means we will have missing values for soil variables for all observations of 2014 and also for some site years that were not in the soil dataset. Since the soil-related variables should be kind of same for the same site, we will impute the missing observations.

```{r}
train_model_dataset <- train_model_dataset %>%
  mutate(
    soilpH_missing = if_else(is.na(soilpH), 1, 0),
    om_pct_missing = if_else(is.na(om_pct), 1, 0),
    soilk_ppm_missing = if_else(is.na(soilk_ppm), 1, 0),
    soilp_ppm_missing = if_else(is.na(soilp_ppm), 1, 0)
  )
```

Mean of soil variables for unique sites


```{r}
site_soil_means <- train_soil_dataset %>%
  group_by(site) %>%
  summarise(
    soilpH_site_mean = mean(soilpH, na.rm = TRUE),
    om_pct_site_mean = mean(om_pct, na.rm = TRUE),
    soilk_ppm_site_mean = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm_site_mean = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )
```

For sites with no observations, we will replace the missing values with the overall median.

```{r}
train_model_dataset <- train_model_dataset %>%
  left_join(site_soil_means, by = "site") %>%
  mutate(
    soilpH = coalesce(soilpH, soilpH_site_mean, median(train_soil_dataset$soilpH, na.rm = TRUE)),
    om_pct = coalesce(om_pct, om_pct_site_mean, median(train_soil_dataset$om_pct, na.rm = TRUE)),
    soilk_ppm = coalesce(soilk_ppm, soilk_ppm_site_mean, median(train_soil_dataset$soilk_ppm, na.rm = TRUE)),
    soilp_ppm = coalesce(soilp_ppm, soilp_ppm_site_mean, median(train_soil_dataset$soilp_ppm, na.rm = TRUE))
  ) %>%
  select(-ends_with("_site_mean"))

colSums(is.na(train_model_dataset))
```

```{r}
glimpse(train_model_dataset)
names(train_model_dataset)
```

```{r}
train_model_dataset %>%
  summarise(
    n_sites = n_distinct(site),
    n_site_original = n_distinct(site_original),
    n_hybrids = n_distinct(hybrid),
    n_previous_crop = n_distinct(previous_crop)
  )
```
```{r}
train_model_dataset <- train_model_dataset %>%
  separate(hybrid, into = c("parent1", "parent2"), sep = "/", remove = FALSE)
```


# E. Modeling strategies  
You are asked to fully train two different machine learning models:  
  - XGboost (everyone)  


1. Pre-processing

```{r}
library(tidymodels)
```


a. Splitting

```{r}
set.seed(931735)
train_data <- train_model_dataset %>%
  filter(year <= 2022)

valid_data <- train_model_dataset %>%
  filter(year == 2023)
```


```{r}
data_split <- make_splits(
  list(analysis = train_data, assessment = valid_data), data = train_model_dataset
)

data_split
```

Distribution of outcome variable (yield_mg_ha)

```{r}
ggplot() +
  geom_density(data = train_data, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = valid_data, 
               aes(x = yield_mg_ha),
               color = "blue")+
  labs(x = "Yield (mg/ha)",
       y = "Density",
       title = "Distribution of yield in training and test data")
```

b. Recipe
```{r}
# Create recipe for data preprocessing
weather_recipe <- recipe(yield_mg_ha ~ ., data = train_data) %>% 
  step_rm(n_plots, year, site, hybrid, soilpH_missing,  om_pct_missing,soilk_ppm_missing, soilp_ppm_missing, parent1, parent2, site_original, previous_crop)
    
weather_recipe
```


```{r}
# Prep the recipe to estimate any required statistics
weather_prep <- weather_recipe %>% 
  prep()

# Examine preprocessing steps
weather_prep
```

```{r}
xgb_spec <- #Specifying XgBoost as our model type, asking to tune the hyperparameters
  boost_tree(
   # Total number of boosting iterations: tune() optimizes number of tree
    trees = tune(),
         # Maximum depth of each tree
    tree_depth = tune(),
             # Minimum samples required to split a node
    min_n = tune(),
        # Step size shrinkage for each boosting step
    learn_rate = tune()
    ) %>%
        #specify engine 
  set_engine("xgboost") %>%
        # Set to mode
  set_mode("regression")

xgb_spec

```

```{r}
set.seed(235) #34549
resampling_foldcv <- vfold_cv(train_data, # Create 10-fold cross-validation resampling object from training data
                              v = 10)

resampling_foldcv
resampling_foldcv$splits[[1]]
```

```{r}
set.seed(123)
xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 50 #It will create 100 combinations of these hyperparameters
  
)

xgb_grid

```


```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)+
  +
  labs(title = "Hyperparameter grid")
```

```{r}
set.seed(76544)
registerDoParallel(cores = parallel::detectCores() - 1)

xgb_res <- tune_race_anova(object = xgb_spec,
                      preprocessor = weather_recipe,
                      resamples = resampling_foldcv,
                      grid = xgb_grid,
                      control = control_race(save_pred = TRUE))
stopImplicitCluster()

beepr::beep()
xgb_res$.metrics[[2]]
```

```{r}
plot_race(xgb_res)
```


```{r}
# Based on lowest RMSE
best_rmse <- xgb_res %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```

```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_res %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

```{r}
# Based on greatest R2
best_r2 <- xgb_res %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```


```{r}
# Based on lowest R2 within 1% loss
best_r2_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_res %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error) %>%
  dplyr::select(source, everything())
```

```{r}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```


```{r}
set.seed(10)

final_fit <- last_fit(final_spec,
                      preprocessor = weather_recipe, split = data_split)

test_predictions <- final_fit %>% collect_predictions()

test_predictions
```


```{r}
final_fit %>%
  collect_metrics()
```


```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, 
                 train_data)) %>%
  augment(new_data = bake(weather_prep, 
                          train_data)) %>% 
  rmse(yield_mg_ha, .pred) %>%
  bind_rows(
    
    
# R2
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, 
                  train_data)) %>%
  augment(new_data = bake(weather_prep, 
                          train_data)) %>% 
  rsq(yield_mg_ha, .pred))
```

```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(20, 40)) +
  scale_y_continuous(limits = c(20, 40)) 
```

```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
         data = bake(weather_prep, train_data)) %>% #There little change in variable improtance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```


  - **One more model of your choice not used in class**.  

You can find a list of all tydymodels supported models here:  
https://www.tidymodels.org/find/parsnip/  

# F. Training strategies  
For both XGboost and your model of choice, the entire training process is up to you.  
That means you decide things like  
  - Data split proportions  
  - Pre-processing steps  
  - Hyper-parameters to fine tune  
  - Search strategy  
  - Resampling strategy  
  - Which metric to use to select best model  
  - Which type of model selection to use (e.g., best model, best within 1 sd, etc.)  
  
# G. Communicating your results  
You will be required to create a **shiny app** to communicate your results.  
Select one of your final models (XGboost or the other one), and display its results in the app.
While the shiny app design is up to you, it should contain:  
  - At least one EDA on yield based on the training data  
  - EDA on at least 1 predictor variable of each group (soil, weather)  
  - At least one interactive component  
  - A plot showing the most important variables in your model  
  - A plot for predicted vs observed with R2 and RMSE on your testing set (not the one I provide)  

# H. Troubleshooting  
You will for sure run into code issues.  
This is common and expected, and part of the learning process.  

While this is an individual project, I do encourage all students to help each other, especially as you will likely run into similar problems.  

For that to work, we will be using **GitHub** to ask and answer questions.  

ALL QUESTIONS should be asked in our course GitHub page (https://github.com/leombastos/2026_dsa) under "Issues". **Please do not use email for asking questions**.

Make sure to **"Watch"** the repository so you get notified when someone posts on Issues.  

> I anticipate all of you will have multiple questions. The goal with using GitHub is that you can help each other. You will be graded for participation both in asking questions on GitHub and also helping others with their questions.  

With that, when you have issues running code, here are a few resources you can use, in chronological order:  

- **Yourself**: Read the error message, see if you can interpret it and understand what is going on. A message like "Error: object yield could not be found" is self-explanatory.    
- **Google**: Sometimes just copying an error message and pasting on Google can help you find posts with the answer.  
- **Peers**: ask your classmates using GitHub.  
- **Me**: after you have gone through all the sources above without success, I will certainly be glad to assist you. I want to be the last resource you use because that's how it will be after our class is finished: I will be available to assist you in anything R-related in your career, but you will also need to attempt solving them before you reach out.  

# I. Turning it in  
You are required to turn in the following:  
  - The github repository URL containing ALL the code for the ML training process  
  - The completed `testing_submission` file with your predictions for the test data that I shared with you. **NOTE: only fill in the NA cells, DO NOT CHANGE anything else! If you send this file with different column names, etc, you WILL LOSE POINTS!**  
  - The URL for the published shiny app  
  - A powerpoint slide named "2026dsa_teamX.ppt" where X is the number of your team. The deck should contain a single slide showing 
  
Submit the github URL, the `testing_submisison` csv, and the shiny app URL by **April 30th** 11:59 pm on the following link:  
https://forms.gle/Es34gk74DHBBSEoa8 
  
# J. Grading  
You will be graded based on:  
  - following general workflows we did in class for all steps (open-source data retrieval and processing, feature engineering, ML model training, etc.)  
  - following reproducible standards like sensible folder and file naming, script organization, script comments, etc.  
  - collaborating with your partner in GitHub  
  - submitting all three pieces of information (2 URLs, 2 files)  
  - submitting on time  
  - having a shiny dashboard with the minimum requirements listed above  

# K. Team work in GitHub  
Whether you are working with your future-self or as duos, make sure to stage, commit, and push after finishing each of the sub-sections above. When committing, write commit messages that are short and descriptive (e.g., finished wrangling).  

If you are working in duos, make sure to split the workload. I will check your collaborative work through the commit history, and if one student has contributed significantly more than the other, than that will impact grades.  

**Tip 1**: to avoid merge conflicts, make sure to **pull** first, and then start working locally. That will ensure that any changes made by your partner will be "downloaded" before you make changes to the files locally.  

**Tip 2**: make use of the Issues on this repository to set up to-do lists and assign tasks to different people. You can also use each issue/task to discuss how things should be run and get to an agreement.  

# L. Extra  
Once each team submits their predictions on my test set, I will compare their predictions with the observed values (that you don't have).  

The team that gets the highest **R2** or lowest **RMSE** will receive an award and bragging rights.  

